---
title: "Simulation-Based Power Analysis"
subtitle: "Part 6: Fully crossed design and by-item varying intercepts"
format:
  revealjs:
    theme: serif
    css: custom.css
    slide-number: true
    chalkboard: true
    code-fold: false
    code-overflow: wrap
    scrollable: true
---

## Part 6: Fully crossed design and by-item varying intercepts

```{r}
#| echo: false
#| eval: true

# load packages
library(tictoc)
library(MASS)
library(lme4)
library(lmerTest)
library(ggplot2)

options(scipen = 999) # prevents scientific notation
set.seed(42) # set seed for reproducibility
```

Why by item varying intercepts?

::: {incremental}
- Words come from a pool of possible items and come with their own variability.
- Each item/word will have its own baseline response time, some will be responded faster, and some slower.
- To simulate this variability we will generate by-item intercept adjustments.
:::

## Part 6: Fully crossed design and by-item varying intercepts

::: {incremental}
- Additionally, in this example each word (item) is seen by every subjects, this is called fully crossed design.
- These data are not independent because responses to the same item are likely to be correlated.
:::

## Part 6: Fully crossed design and by-item varying intercepts

::: incremental
- We will generate by-item intercept adjustments.
- We will account for this variability/dependency in response times by including the following term: (1\|item).
- Historically, in psycholinguistic studies, two ANOVAs would be run: one averaged over subjects (F1) and one averaged over items (F2). Linear mixed models allow us to include both sources of variability in a single model.
:::

## Part 6: Fully crossed design and by-item varying intercepts

**Generate fully crossed design:**

```{r}
#| echo: true
#| eval: true
#| output-location: fragment
#| code-line-numbers: "|4-5"

n = 20  # subjects
k = 20  # 20 items per condition

# Create fully crossed design: each subject sees all items
sim_dat <- expand.grid(item = 1:(k*2), subj = 1:n)

# Assign items to conditions (half low, half high frequency)
sim_dat$cond.cod <- rep(c(-0.5, 0.5), each = k)

sim_dat
```

## Part 6: Fully crossed design and by-item varying intercepts

**Generate fully crossed design:**

```{r}
#| echo: true
#| eval: true
#| output-location: fragment

# Sanity check:
head(xtabs(~subj + item, data = sim_dat))
```

## Part 6: Fully crossed design and by-item varying intercepts

**Generate by-item intercept adjustments:**

```{r}
#| echo: true
#| eval: true
#| code-line-numbers: "|1|3-4|6"
#| output-location: fragment

sigma_w = 100   # SD of item random INTERCEPTS
  
# Generate item random effects:
w <- rnorm(n = length(unique(sim_dat$item)), mean = 0, sd = sigma_w)

w
```

## Part 6: Fully crossed design and by-item varying intercepts

**What is the new term here?**

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|3"

rt[i] <- rnorm(1, b0 +                      # grand mean (intercept)
                 u[sim_dat[i,]$subj,1] +    # subject random intercept
                 w[sim_dat[i,]$item] +      # item random intercept
                 (b1 +                      # fixed effect of frequency
                 u[sim_dat[i,]$subj,2]) * sim_dat[i,]$cond.cod, # subject slope adjustment
                 sigma)                      # residual error

```

## Part 6: Fully crossed design and by-item varying intercepts

**What is the new term here?**

```{r}
#| echo: true
#| eval: false

rt[i] <- rnorm(1, b0 +                      # grand mean (intercept)
                 u[sim_dat[i,]$subj,1] +    # subject random intercept
                 w[sim_dat[i,]$item] +      # item random intercept
                 (b1 +                      # fixed effect of frequency
                 u[sim_dat[i,]$subj,2]) * sim_dat[i,]$cond.cod, # subject slope adjustment
                 sigma)                      # residual error

```

$$
rt_{ij} = \beta_0 + u_{0i} + w_{0j} + (\beta_1 + u_{1i}) \times cond.cod_{ij} + \varepsilon_{ij}
$$ \

- $w_{0j} \sim Normal(0, 100)$; by-subject adjustments to the grand mean


## Part 6: Fully crossed design and by-item varying intercepts

**Putting it all together:**

```{r}
#| echo: true
#| eval: true

gendat_ldt_crossed <- function(n = 20,          # subjects
                               k = 20,          # 20 per condition
                               b0 = 725,        # grand mean
                               b1 = 50,         # log-scale frequency effect
                               sigma_u0 = 150,  # SD of subject random intercepts
                               sigma_u1 = 20,   # SD of subject random slopes
                               rho = -0.5,      # correlation between subject intercepts and slopes
                               sigma_w = 100,   # SD of item random INTERCEPTS only
                               sigma = 200){    # residual SD

  # Create fully crossed design: each subject sees all items
  sim_dat <- expand.grid(item = 1:(k*2), subj = 1:n)

  # Assign items to conditions (half low, half high frequency)
  sim_dat$cond.cod <- rep(c(-0.5, 0.5), each = k)

  # Build variance-covariance matrix for subjects from individual parameters
  Sigma_u <- matrix(c(sigma_u0^2, rho * sigma_u0 * sigma_u1,
                      rho * sigma_u0 * sigma_u1, sigma_u1^2), nrow = 2)

  # Subject random effects (intercepts AND slopes):
  u <- mvrnorm(n = length(unique(sim_dat$subj)),
               mu = c(0,0), # random intercept and slope
               Sigma = Sigma_u)

  # Item random effects (intercepts ONLY):
  w <- rnorm(n = length(unique(sim_dat$item)), mean = 0, sd = sigma_w)

  # generate data row by row:
  N <- dim(sim_dat)[1]
  rt <- rep(NA, N)
  for(i in 1:N){
    rt[i] <- rnorm(1, b0 +                      # grand mean
                        u[sim_dat[i,]$subj,1] + # subject random intercept
                        w[sim_dat[i,]$item] +   # item random intercept
                      (b1 + 
                         u[sim_dat[i,]$subj,2]) * sim_dat[i,]$cond.cod,
                    sigma)                     # residual error
  }
  sim_dat$rt <- rt
  sim_dat$subj <- factor(sim_dat$subj)
  sim_dat$item <- factor(sim_dat$item)
  sim_dat
}
```

## Part 6: Fully crossed design and by-item varying intercepts

**Test the crossed design:**

```{r}
#| echo: true
#| eval: true
#| output-location: fragment

sim_dat <- gendat_ldt_crossed(n = 20, 
                              k = 20,
                              b0 = 725,
                              b1 = 50,
                              sigma_u0 = 150,  # SD subject intercepts
                              sigma_u1 = 20,   # SD subject slopes
                              rho = -0.5,      # correlation
                              sigma_w = 100,   # SD item intercepts
                              sigma = 200)     # residual SD

# Check the structure
dim(sim_dat)  # Should be 20 subjects × 40 items = 800 observations
```

## Part 6: Fully crossed design and by-item varying intercepts

**Test the crossed design:**

```{r}
#| echo: true
#| eval: true
#| output-location: fragment

head(sim_dat, 41)
```

## Part 6: Fully crossed design and by-item varying intercepts

**Verify fully crossed structure**

-   Each subject sees 20 items per condition:

```{r}
#| echo: true
#| eval: true
#| output-location: fragment

xtabs(~subj + cond.cod, data = sim_dat)  
```

## Part 6: Fully crossed design and by-item varying intercepts

**Verify fully crossed structure**

-   Each item is responded by 20 subjects:

```{r}
#| echo: true
#| eval: true
#| output-location: fragment

xtabs(~cond.cod + item, data = sim_dat)  
```

## Part 6: Fully crossed design and by-item varying intercepts

**Verify fully crossed structure**

-   1 observation per subject-item pair:

```{r}
#| echo: true
#| eval: true
#| output-location: fragment

xtabs(~subj + item, data = sim_dat)      
```

## Part 6: Fully crossed design and by-item varying intercepts

-   Do the response times data look ok to you or is there something off?

```{r}
#| echo: true
#| eval: true
#| output-location: fragment

sim_dat <- gendat_ldt_crossed(n = 64, k = 40)
hist(sim_dat$rt, breaks = 30, main = "Histogram of simulated RTs",
     xlab = "Reaction Time (ms)")
```

## In this section we have:

...generated data with correlated by-subject adjustments to intercepts and slopes and by-item adjustments to intercepts:

$$
rt_{ij} = \beta_0 + u_{0i} + w_{0j} + (\beta_1 + u_{1i}) \times cond.cod_{ij} + \varepsilon_{ij}
$$ where $\varepsilon_{ij} \sim Normal(0, \sigma)$ and

$$
\begin{pmatrix} u_0 \\ u_1 \end{pmatrix} \sim MVN \left( \begin{pmatrix} 0 \\ 0 \end{pmatrix}, \Sigma_u \right), \quad
$$

$$
\Sigma_u = \begin{pmatrix}
\sigma_{u0}^2 & \rho_u \sigma_{u0} \sigma_{u1} \\
\rho_u \sigma_{u0} \sigma_{u1} & \sigma_{u1}^2
\end{pmatrix} \quad
$$ $$
\ w_0 \sim N (0, \sigma_w)
$$

## Next step:

-   We will now model log(rt) to ensure positive reaction times and more realistic distribution.

## Preguntas? {background-color="#43464B"}